Investigation of thermal reactionomes reveals stress tolerance vs. resistance in

### 2014-08-22 doc/ApTranscriptome_manuscript.Rmd
Added Significance Statement

Thermal tolerance is a key parameter underlying species distributions, and varies considerably within species. However, little is known about the genetic basis of thermal tolerance or how it has evolved. In recent years, whole organism gene expression (e.g. transcriptomic) studies have been used to gain greater insight on these questions, but these studies are limited to finding genes with large effects in a few conditions. Here, we present a method to characterize the reaction norms of gene expression across an ecologically-relevant thermal gradient (e.g. the thermal reactionome) for all genes in the transcriptomes of two closely related ant species. We find that a large fraction of the transcriptome is thermally-responsive. The warmer-climate species down-regulates more genes in response to thermal extremes than the colder-climate species. These differences in expression patterns are indicative of divergent strategies in these two species; stress tolerance in the warmer-climate species and stress-resistance in the colder-climate species.

>>>>

Successfully coping with progressive changes in climate will require both short-term physiological acclimation and longer-term adaptive changes. To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found surprisingly divergent transcriptomic responses in the two species. *A. picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a more robust heat shock response. In contrast, *A. carolinensis* reacted to thermal stress to a greater extent through downregulation, reducing metabolic gene expression and requiring more extreme temperatures to trigger a protective response. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely-encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatic zones.
